---
title: "Introduction to ggplot: course code"
author: "Dr. Laurie Baker"
date: "21/04/2020"
output: 
  html_document:
    number_sections: true
    toc: yes
    toc_depth: '2'
    toc_float: yes
---

# Getting started

## Installing and loading packages

```{r libraries, echo=TRUE, message=TRUE, warning=TRUE}
# install.packages("tidyverse")
# install.packages("gapminder")

library(tidyverse)
library(gapminder)

```

## Cleaning and prepping the data

### This is messy data

```{r tidy-messy1, results='asis', echo=FALSE}

pop_simple <- gapminder %>% 
  filter(
    country %in% c("Syria", "Rwanda", "Chile"), 
    year >= 1997
  ) %>% 
  select(country, year, pop) %>% 
  mutate(pop = pop / 10^6)

messy_pop <- pop_simple %>% spread(year, pop)

knitr::kable(messy_pop, format = 'html')

```
## Cleaning and prepping the data


```{r tidy-tidy1, result='asis'}

tidy_pop <- gather(messy_pop, 'year', 'pop', -country)
tidy_pop %>%
  head(4) 

```


## Building a plot

### Calling `ggplot()`

```{r first-plot1a, ref.label='first-plot1a', out.width="100%"}

ggplot(tidy_pop)

```

### Adding aesthetics

```{r first-plot1b, ref.label='first-plot1b', out.width="100%"}

ggplot(tidy_pop,
  aes(x = year, #<<
      y = pop) #<<
  )
```

### Choosing the geometry (geom)

```{r first-plot1c, ref.label='first-plot1c', out.width="100%"}

ggplot(tidy_pop,
  aes(x = year,
      y = pop)
  ) +
  geom_point() 

```

### Adding a third variable, country, as a colour.

```{r first-plot1d, out.width = "100%"}

ggplot(tidy_pop,
  aes(x = year,
      y = pop,
      color = country) 
  ) + 
  geom_point()

```

### Adding a line to the plot

```{r first-plot1e, out.width = "100%"}

ggplot(tidy_pop,
  aes(x = year,
      y = pop,
      color = country) 
  ) +
  geom_point() +
  geom_line()

```

### Adding a line to the plot, group by country.

```{r first-plot1f, out.width = "100%"}

ggplot(tidy_pop,
  aes(x = year,
      y = pop,
      color = country) 
  ) +
  geom_point() +
  geom_line(
    aes(group = country)) 

```

### Storing a plot

```{r first-plot1g, out.width = "100%"}

g <- ggplot(tidy_pop, 
  aes(x = year,
      y = pop,
      color = country) 
  ) +
  geom_point() +
  geom_line(
    aes(group = country))

g 


```

### Specifying the aesthetics for each geometry.

```{r}

g <- ggplot() +
  geom_point(
    data = tidy_pop, 
    aes(x = year, 
      y = pop, 
      color = country) 
  ) +
  geom_line(
    data = tidy_pop, 
    aes(x = year, 
      y = pop, 
      color = country, 
      group = country) 
    )

g

```

### Adding more data. Creating `tidy_pop2`

```{r}

tidy_pop2 <- left_join(tidy_pop, select(gapminder, country, continent))

g <- ggplot(tidy_pop2) +
  aes(x = year,
      y = pop,
      color = country) +
  geom_point() +
  geom_line(aes(group = country))

```

### Using Facets

```{r geom_facet}

g + facet_wrap(~ country)

```

### Using facet_grid, two variables.

```{r geom_grid}
g + facet_grid(continent ~ country)
```

### Adding labels

```{r labs-ex}

(g <- g + labs(x = "Year", y = "Population (millions)"))

```

### Showing coordinates

```{r adding coordinates}

g + coord_flip()

```

### Changing the color scale

```{r}

g + scale_color_manual(values = c("peru", "pink", "plum"))

```

### Changing the scale of the y

```{r}

g + scale_y_log10()

```

### Changing the scale of the x

```{r}

g + scale_x_discrete(labels = c("MCMXCVII", "MMII", "MMVII"))

```

### Changing the theme

```{r theme ex1}

g + theme_bw()

```

### Changing the theme (fonts and colours) 

```{r theme ex2}

g + theme_bw() + theme(text = element_text(colour = "hotpink", size = 20))

```

### Setting an overall theme:

```{r Setting a theme}

my_theme <- theme_bw() +
  theme(
    text = element_text(family = "Palatino", size = 12),
    panel.border = element_rect(colour = 'grey80'), 
    panel.grid.minor = element_blank()
  )

theme_set(my_theme)

```


### Making the plot with the new theme

```{r}

theme_set(my_theme)
g 

```

### Changing the position of the legend

```{r}

theme_set(my_theme)
g + theme(legend.position = 'bottom')

```

