---
title: "Introduction to R Part II: Exercises"
author: "Dr. Laurie Baker"
date: "15/04/2020"
output: 
  html_document:
    number_sections: true
    toc: yes
    toc_depth: '1'
    toc_float: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading in the relevant packages and the data
 
```{r data and libraries}

library(tidyverse)

### If your data folder is outside your exercises folder:

my_file <- file.path("../data", "gapminder.csv")

### If your data folder is in your exercises folder:

#my_file <- file.path("data", "gapminder.csv")

gapminder <- read_csv(file = my_file)

```

# Subsetting and summarising using dplyr:

`dplyr`, has made a lot of data manipulation easier and clearer using **verbs** to filter and select different elements.



  * `select()` subsets columns based on their names.



  * `filter()` subsets rows based on their values.



  * `summarise()` calculates summary statistics.



  * `group_by()` groups variable for summarising.



  * `mutate()` adds new columns that are functions of existing variables. 



These verbs can be combined in powerful ways to do some really interesting data manipulation tasks! 

## select

```{r Select columns lifeExp and pop using select()}

select(gapminder, lifeExp, pop)

```



## The pipe operator

* `%>%` = **"and then"**

```{r Select the columns lifeExp and country using select()}

gapminder %>%
  select(lifeExp, country)

```


## Assign your output to a new data frame


```{r Create a new data frame lifeExp_by_country with the columns lifeExp and country}

lifeExp_by_country <- gapminder %>%
  select(lifeExp, country)

head(lifeExp_by_country)

```


class: inverse, center, middle

# Exercises Part 1



## 1. Run the following line of code, what does the minus do?

```{r Using - to remove columns, eval = FALSE}

gapminder %>%
  select(-c(lifeExp, country))

```


## 2. Select the columns `country`, `continent` and `gdpPercap` from the data frame. 

```{r Select the columns country, continent and gdpPercap using select, eval = FALSE}

gapminder %>%
  select()

```


## 3. Write 2 ways to select all the columns except for year.

**Option 1**


```{r Write code for two ways to select all columns but year, eval = FALSE}
gapminder %>%
  select()

```



**Option 2**

```{r Write code for two ways to select all columns but year 2, eval = FALSE}

gapminder %>%
  select() 

```



## filter

* `filter`: subsetting rows


## Filter using double "="

* Filter a particular country:

```{r Use filter to pick out a particular country}

gapminder %>%
  filter(country == "Yemen, Rep.") %>%
  head()

```



## Filter rows from a set of matches


```{r Use filter to filter row from a set of countries}

gapminder %>%
  filter(country %in% c("Jordan", "Oman",
                        "Syria", "Yemen, Rep."))

```


## Combining multiple filters

* You can add multiple filters with a comma.

```{r Add multiple filters with a comma}

gapminder %>%
  filter(country == "Yemen, Rep.", year >= 1960 & year <= 1985)

```


class: inverse, center, middle

# Exercises Part 2


## 1. What do these lines of code filter the data for?

```{r What does this code filter for, eval = FALSE}

gapminder %>%
  filter(continent == "Europe", lifeExp > 70)

```


## 2. Filter the data for countries in "Asia" where the "lifeExp" was below 35  

```{r Filter the data for countries in Asia where lifeExp is less than 35, eval = FALSE}

gapminder %>%
  filter(continent == BLANK, lifeExp BLANK)
  
```


## 3. Filter the data for observations where the gdpPercap was equal to 1000 or less. 

```{r Filter the data so that you only get entries where gdpPercap is equal to 1000 or less, eval = FALSE}

gapminder %>%
  filter(gdpPercap BLANK)
  
```

**BONUS**

## 4. Filter using `%in%` for countries "Chile", "Argentina", "Uruguay", and "Peru" and year >= to 1992. 

```{r Filter the data for Chile Argentina Uruguay and Peru for years greater than or equal to 1992, eval = FALSE}

gapminder %>%
  filter(country %in% BLANK & year BLANK)


```

**BONUS**

## 5. Filter the data using `!=` to include the data from all continents apart from Europe. 

```{r Filter the data to inclue data from all continents apart from Europe, eval = FALSE}

gapminder %>%
  filter(continent BLANK)

```


## summarise()



* `summarise()` uses existing R functions to calculate summary statistics. 



* mean lifeExp for all countries:



```{r Calculate the mean_lifeExp using summarise}

(lifeExp_stats <- gapminder %>%
                  summarise(
                    mean_lifeExp = mean(lifeExp) 
                    )
 )

```


## summarise() multiple summary statistics

* Calculate multiple summary statistics by separating using ","

```{r Use summarise to calculate mean_lifeExp, min_lifeExp, and max_lifeExp}

(lifeExp_stats <- gapminder %>%
                  summarise(
                    mean_lifeExp = mean(lifeExp), # mean 
                    min_lifeExp = min(lifeExp), # min 
                    max_lifeExp = max(lifeExp)) # max 
                    ) 


```



## group_by()



* `group_by()` used to group variables. 


* summarise min, mean, and max `lifeExp` per country



```{r Use group_by to group by country and then summarise}

(lifeExp_stats_country <- gapminder %>%
                             group_by(country) %>% 
                             summarise(
                                mean_lifeExp = mean(lifeExp),
                                min_lifeExp = min(lifeExp), 
                                max_lifeExp = max(lifeExp)
                                ))

```



class: inverse, center, middle

# Exercises Part 3


## 1. What does the following bit of code do? 

```{r What does the following group_by summarise code do, eval = FALSE}

gapminder %>%
      group_by(continent, year) %>%
      summarise(mean_gdpPercap = mean(gdpPercap))


```


## 2. Group the data by country and summarise the minimum and maximum population sizes. 


```{r Group the data by country and create two new variables which summarise the minimum and maximum pop, eval = FALSE}

gapminder %>%
  group_by(BLANK) %>%
  summarise(min_pop = BLANK,
            max_pop = BLANK) %>%
            head(8)
          
```


## 3. Group the data by continent and year. Summarise the max and min population.

```{r Summarise maximum and Minimum population by continent and year, eval = FALSE}

gapminder %>%
  group_by(BLANK, BLANK) %>%
  summarise(min_pop = min(pop),
            max_pop = max(pop)) %>%
            head(8)
          
```




# The pipe function %>%

## The pipe function in action!


```{r Yemen Life Expectancy pre 1980 3}


yemen_pre1980_mean_lifeExp <- gapminder %>%
    filter(country == "Yemen, Rep.", year <1980) %>% 
    select(lifeExp) %>% 
    summarise(mean_lifeExp = mean(lifeExp)) 


yemen_pre1980_mean_lifeExp

```



## Looking at a slice()


`slice()` chooses rows by their position within the group, e.g. minimum `lifeExp`.

```{r Using group_by and slice: min lifeExp}

gapminder %>%
  group_by(year) %>%
  slice(which.min(lifeExp)) 

```


## Looking at a slice()

* Which country had the highest life Expectancy in each year?

```{r Using group_by and slice: max lifeExp}

gapminder %>%
  group_by(year) %>%
  slice(which.max(lifeExp))

```


## Mutate

  * `mutate()` adds new columns that are functions of existing variables. 
  


```{r Creating a new variable gdp}

(gapminder <- gapminder %>%
    mutate(gdp = gdpPercap*pop)) 



```



  * gdp is useful... but it's also a huge number! Which makes it difficult to compare. 
  


# Joining data frames: when one data frame is not enough

* Our data is often spread out over several data frames!

* We can **join** these data frames together using a variety of `join` functions from the `dplyr` package.

Let's walk through the different types of joins using a simple example. 


## Joining data

**Person Table**

```{r}

(person_table <- data.frame(Person_ID = c("Person1", "Person2"), 
                           Name = c("Jane Doe", "John Smith"), 
                           Job_ID = c("Job_1", "NA")))
```



**Job Table**  
   
```{r}                           
(job_table <- data.frame(Job_ID = c("Job_1", "Job_2"), 
                        Job_Name = c("Programmer", "Statistician")))
```


## Inner join:


```{r}
person_table
```

```{r}
job_table
```



* `inner_join:` only rows where there is a match. 

```{r Inner join example, warning = FALSE}

inner_join(x = person_table, y = job_table, by = "Job_ID")

```


## Left join:


```{r}
person_table
```

```{r}
job_table
```



* `left_join:` All rows on the left side of join. 



* Only rows from the right side where there's a match on the left. 



```{r Left join example, warning = FALSE}


left_join(x = person_table, y = job_table, by = "Job_ID")

```


## Right join:

```{r}
person_table
```

```{r}
job_table
```



* `right_join:` Rows from the left side only where there is a match on the right. 



* All rows from the right side of the join.

```{r Right join example, warning = FALSE}

right_join(x = person_table, y = job_table, by = "Job_ID")

```



## Full join

```{r}
person_table
```

```{r}
job_table
```



* `full_join:` All rows from the left.

* All rows from the right, **Joined** where the criteria matches. NAs where no matches.

```{r Full join example, warning = FALSE}

full_join(x = person_table, y = job_table, by = "Job_ID")

```


# Creating a new data frame uk_gdpPercap_df

* Let's look at the per capita GDP in a way that's more meaningful.



* Let's create `gdpPercap_rel`, that is the `gdpPercap` of the country **relative** to the UK's `gdpPercap` of that same year.



* How? Country A: `gdpPercap` divided by UK's `gdpPercap` in the same year.



# Creating a new data frame uk_gdpPercap_df

* Create a new dataframe `uk_gdpPercap_df`

* Filter the rows for `country == "United Kingdom"`.

* Select the columns `gdpPercap` and `year`.

* Rename the variable `gdpPercap` to `uk_gdpPercap`.

* Inspect the data frame.

```{r Make a new data frame uk_gdpPercap_df 5}

uk_gdpPercap_df <- gapminder %>%  
  filter(country == "United Kingdom") %>% 
  select(gdpPercap, year) %>% 
  rename(uk_gdpPercap = gdpPercap)

head(uk_gdpPercap_df) 

```


## Joining `uk_gdpPercap_df` to the `gapminder` data frame

* We want to divide all the other country's `gdpPercap` by the UK `gdpPercap` in that same year.
 

 
* We can match the two data frames using a `left_join` on the common variable, `year`.



```{r Join the gapminder and uk_gdpPercap_df data frames}

gapminder <- left_join(gapminder, uk_gdpPercap_df, by = "year")

names(gapminder)

```

 
## Calculating the relative GDP per capita

```{r Make a new variable gdpPercap_rel}

gapminder <- gapminder %>%
  mutate(gdpPercap_rel = gdpPercap/uk_gdpPercap) 

```

* Doublechecking our calculations... Is the United Kindom relative gdp always 1?


```{r Doublechecking our calculation of gdpPercap_rel worked}

gapminder %>%
  filter(country == "United Kingdom") %>%
  select(gdpPercap_rel) %>%
  head()

```

We can doublecheck that our calculation worked by filtering for the United Kingdom to check that the relative gdp per capita is 1. 

Now that we have the `gdpPercap` and `uk_gdpPercap` matched up, we can calculate the relative GDP per capita `gdpPercap_rel`. 



## How many countries had a smaller gdp per capita than the UK each year?

```{r How many countries had a smaller gdpPercapita}

gapminder %>%
  group_by(year) %>%
  filter(gdpPercap_rel < 1) %>%
  summarise(count = n()) 

```


# Exercises Part 4

## 1. What does the following bit of code do?

```{r 1. Select 2. filter 3. group_by and 4. summarise, eval = FALSE}

gapminder %>%
  select(country, gdpPercap_rel) %>%
  filter(country %in% c("Argentina", "Chile", "Peru", "Brazil")) %>%
  group_by(country) %>%
  summarise(
            max_gdp = max(gdpPercap_rel), 
            min_gdp = min(gdpPercap_rel), 
            mean_gdp = mean(gdpPercap_rel)
            )

```


## 2. How many countries had a higher relative gdp per capita than the United Kindom per year?

```{r How many countries had a higher relative gdp per capita, eval = FALSE}

gapminder %>%
  group_by() %>%
  filter(gdpPercap_rel BLANK) %>%
  summarise(count = n())

```


## 3. **Which countries** had a higher relative gdp per capita than the UK in 2007? Fill in the blanks

```{r Which countries had a higher relative gdp per capita than the UK, eval = FALSE}

gapminder %>%
  filter(BLANK, year == 2007) %>%
  select(country) %>%
  unique()

```


# Answers to our poll 


## Which country had a higher infant mortality rate in 2007? 

* Sri Lanka or Turkey

```{r Turkey or Sri Lanka: who has a higher infant mortality rate in 2007, eval = TRUE}

gapminder %>%
  filter(year == 2007, country %in% c("Sri Lanka", "Turkey")) %>%
  select(country, infant_mortality)

```

* **Turkey**, difference = 9.5.

## Which country had a higher infant mortality rate in 2007? 

* Poland or Malaysia

```{r Poland or Malaysia: who has a higher infant mortality rate in 2007, eval = TRUE}

gapminder %>%
  filter(year == 2007, country %in% c("Poland", "Malaysia")) %>%
  select(country, infant_mortality)


```



* **Poland**, difference = 0.9


## Which country had a higher infant mortality rate in 2007?

* Pakistan or Vietnam


```{r Pakistan or Vietnam: who has a higher infant mortality rate in 2007, eval = TRUE}

gapminder %>%
  filter(year == 2007, country %in% c("Pakistan", "Vietnam")) %>%
  select(country, infant_mortality)

```



* **Pakistan**, difference = 56


## Which country had a higher life Expectancy in 2007? 

* South Africa or Yemen

```{r South Africa or Yemen: who has the highest life Expectancy}

gapminder %>%
  filter(year == 2007, country %in% c("South Africa", "Yemen, Rep.")) %>%
  select(country, lifeExp)

```


* **Yemen**, difference ~ 13.4 years


## Which country had a higher life Expectancy in 2007? 

* Chile or Hungary

```{r Chile or Hungary: who has the highest life Expectancy, eval = TRUE}

gapminder %>%
  filter(year == 2007, country %in% c("Chile", "Hungary")) %>%
  select(country, lifeExp)

```



* **Chile**, difference ~ 5.2 years


## Which country do you think had the highest gdpPercap in 2007?

* Switzerland or Kuwait

```{r Switzerland or Kuwait: who has the highest gdpPercap, eval = TRUE}

gapminder %>%
  filter(year == 2007, country %in% c("Switzerland", "Kuwait")) %>%
  select(country, gdpPercap)

```




* **Kuwait**, difference ~ 10,000 USD, about 20-25% more



## Which country do you think had the highest gdpPercap in 2007?


* Colombia or Nepal


```{r Colombia or Nepal: who has the highest gdpPercap, eval = TRUE}

gapminder %>%
  filter(year == 2007, country %in% c("Colombia", "Nepal")) %>%
  select(country, gdpPercap)

```



* **Colombia**, difference ~ 5,915, more than 3x

