---
title: "Introduction to R Part I: Exercises"
author: "Dr. Laurie Baker"
date: "15/04/2020"
output: 
  html_document:
    number_sections: true
    toc: yes
    toc_depth: '1'
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Basics

## R Notebooks

* Notebooks combine code and descriptions!

```{r Print a statement}

print('code goes here!')

```

### Adding Comments


* Comments are added using `#`.


```{r Print a statement 1}

## print() is a function to print a statement

print('code goes here!')

```

## Objects

* Let's start by making an assignment and inspecting the object we created.

```{r First R Object}

x <- 10*5

x

```

### Names

```{r Names that work for objects}

this_works <- 10*5

this_works

```

* This doesn't work...

```{r Names that do not work for objects, eval = FALSE}

# this_doesn't_work <- 10*5

```

### Using Tab Completion to Complete Object Names

* Make a new object 

```{r Make a new object with a long name}

a_very_long_name <- 7^2 

```

* Start by typing the first few letters of `a_very...` in the code chunk below and type `TAB` to complete the name. 

```{r Using Tab to Test out the Completion Facility in RStudio, eval = FALSE}

a_very

```

### R is case-sensitive and dislikes typos

* Let's try inspecting the object again.

```{r R is case-sensitive and does not like typos, eval = FALSE}
# What happens if you run:

a_vry_long_name

A_very_long_name

```

## Vectors


* A `vector` is a 1-dimensional ordered collection of elements where the elements are **all of the same type.** 

* It is the fundamental data structure in R with a lot of useful properties. 
### Creating a vector with `c()`

* Let's make a new vector called `character_vector` using the function `c()` which can be used to **c()**ombine elements.  

```{r Define a character vector}

## Defining the character vector:

character_vector <- c("ET", "phone", "home", "ET", "phone", "home")

```

### Checking the structure of the vector using `str()`

```{r Checking the structure of the character vector}

str(character_vector)

```

### Check the length of a vector using `length()`

```{r Checking the length of the character vector}

length(character_vector)

```

### Extract multiple consecutive elements using `:`. 

* We can extract an element from a vector by referencing its position. 


```{r Take the 3rd and 4th and 5th elements}

character_vector[3:5]

```


### Replace elements using `<-`

Try replacing the 4th element with your name:

```{r Replace an element}

character_vector[4] <- "Laurie"

character_vector
```

### Define a numeric vector

* The same method used to extract information works for any type of vector. 

* Let's create a new `numeric_vector` containing the numbers 1, 2, 3, 4, and 5. 


```{r Define a numeric vector}

numeric_vector <- c(1:5) 

```

### Check the structure using `str()`

```{r Check the structure of a numeric vector}

str(numeric_vector)

```


### Extract the first two elements

```{r Take the first two elements from a numeric vector}

numeric_vector[1:2]

```


### Extract non-consecutive elements using `c()`

Trying uncommenting and running the line below:

```{r We always need to use c() to specify multiple non-consecutive positions}

# numeric_vector[1,3]

```


Note that if we want, we can only select the 1 and 3 or 1, 3, and 4 elements using `c()`. 

```{r Use c() to specify non-consecutive positions}

numeric_vector[c(1,3:4)]

```


### A dynamically typed language

* R will try to be smart and figure out your data type for you

```{r Make a second numeric vector}

numeric_vector2 <- c(1.1,3:4)

## Check the structure

str(numeric_vector2)

```

### The `as.` functions: integer to numeric


* The `as.` functions take the form `as.<structure>`.

* We want to make our vector numeric so, `<structure>` =  numeric

```{r Change the structure of a numeric vector using as. functions}

numeric_vector <- as.numeric(numeric_vector)

str(numeric_vector)

```


### The `as.` functions: character to factor

* Example 2: Change character to factor. 

```{r Change the structure of a character vector using as. functions}

character_vector <- as.factor(character_vector)

str(character_vector)


```

## Vectorised Language

* R is a **vectorised language.**

* Operations are applied to each element automatically without the need to loop through!

* Vectorised = Fast! Computers are optimised for these types of calculations. 

* Can mutiply, divide, add, and subtract vectors of the same length or multiples of the same length
### Can multiply and exponentiate vectors

```{r You can multiply and exponentiate vectors}

numeric_vector*3

numeric_vector^2

```

### Divide vectors of the same length

```{r Divide vectors of the same length}

x <- seq(from = 1, to = 20, by = 4)

x

numeric_vector/x

```

### Subtract or add vectors of the same length


```{r You can subtract vectors of the same length}

x - numeric_vector

```

### Many functions are designed for vectors

* **Let's calculate some simple statistics**

```{r}

mean(numeric_vector)

min(numeric_vector)

max(numeric_vector)

range(numeric_vector)

```


# Exercises Part I

1. Take the last two elements of the numeric vector

Hint: use length() to find out the last element

```{r Take the last two elements of the numeric vector, eval = FALSE}

numeric_vector[]

```


2. Take the first and last elements of the character_vector. 

Hint: you can use `length()` to find out how many elements there are in the character vector. 

```{r Take the first and last elements of the character_vector, eval = FALSE}

character_vector[]

```

3. Divide the numeric_vector by 3

```{r Divide the numeric vector by 3, eval = FALSE}

numeric_vector

```

4. Multiply the numeric_vector by the new vector 'y'

```{r Multiply the numeric_vector by the new vector y, eval = FALSE}

y <- c(5:1)



```

5. Why do the following lines of code not work?


```{r Why does dividing numeric_vector by the object w not work, eval = FALSE}

w <- c(1:4)

numeric_vector/w

```


# CRAN, Library, Function, Package

```{r Importing Packages, eval = TRUE}

library(tidyverse)

```

## Functions


Functions follow the form:

* `functionName(argument1 = value1, argument2 = value2, and so on)`

Let's take a look at `seq()`, which makes regular sequences of numbers.

* Type `se` and hit TAB. A pop up shows you possible completions. 

```{r Autocompleting using TAB, eval = FALSE}

se

```

* select `seq` from the drop-down menu or complete the name with more typing. Add the numbers "1,10" and run the function.


```{r Using the function seq}

seq(1,10)

```

### Help files: `seq()`


```{r Helpfile for seq(), eval = FALSE}

?seq

```

### How are function arguments resolved?

What happens if we try:

```{r How are function arguments resolved part 1, eval = T}

seq(10,1)

```


And what about:

```{r How are function arguments resolved part 2, eval = T}

seq(to = 10, from = 1)

```

R resolves **first by name and then by position**

### Printing objects and viewing your workspace

* To store our output and see it in the same line, we can use:

```{r Storing and printing objects using parentheses, eval = TRUE}

(y <- seq(from = 1, to = 10))

```


* We can look at our workspace using `ls()`.  

```{r Use ls() to see what is in your workspace, eval = TRUE}

ls()

```

* Remove the vector name `y`:

```{r Removing vectors using rm(), eval = TRUE}

rm(y)

```

* Remove everything using:

```{r Remove everything using rm() and ls()}

rm(list = ls())

```

# Data frames and tibbles

* Data frames are essentially rectangular, spreadsheet-like data.

* Data frames can hold **variables** of different types.


## Reading in the data

Before you read in a data file you want to ask yourself two questions:

1. What type of file is it?
2. Where is the file stored?

In this case, we are going to read in a `.csv` (comma separated value) file called `gapminder.csv`. 


## Reading in the data using the `read_` functions

The functions for reading in the data take the same basic form
 
```{r Layout of the read functions, eval = TRUE}

my_file <- file.path("data", "gapminder.csv")

gapminder <- read_csv(file = my_file)

```

* First you need to specify the **name of the data frame** you want to store your data in. 
  * my_file

* Then the file name and location
  * read_csv(file = ".csv")
  * read_excel(path = ".xlsx")


## Getting to know the data

There are several tools to get to know our data.

 * `View()`: allows us to view the data frame as a spreadsheet.


 * `nrow()`: tells us the number of rows in our data frame.


 * `names()`: gives us the names of the columns in our data frame.


 * `dim()`: tells us the dimensions of our data frame.


 * `summary()`: give us summary statistics (counts, min, median, mean, max). 


 * `head()`: gives us the first 6 elements of the data.


 * `tail()`: gives us the last 6 elements of the data.


 * `str()`: tells us the variable type (e.g. Factor, num (number), int (integer)). 
 

 * `unique()`: tells us the unique elements of a variable.

## Using head()

Let's take a look at `head` and `View` to inspect the data more closely. 

```{r Look at the top 6 rows using head()}

head(gapminder)

```

---
## Using View()

```{r Look at the data using View(), eval = FALSE}

View(gapminder)

```


From viewing the data we can see that the data contains eight variables: `country`, `continent`, `year`, `lifeExp`, `pop`, `gdpPercap`, `infant_mortality`, and `fertility`.

### Checking the structure of the data using str()

```{r Check the structure of the data using str()}

str(gapminder)

```
### Changing the character data to factors

```{r Changing gapminder characters to factors}

gapminder$country <- as.factor(gapminder$country)
gapminder$continent <- as.factor(gapminder$continent)


str(gapminder)
```

# Exercises Part II

Click "Filter" in the View menu, you can use this similarly to how you would interact with the data in Excel. 

1. Using filter, what was the life expectancy in Rwanda in 1952?

2. Which country had the highest infant mortality rate? What was the year?

**Next**


Run the following lines of code to answer the questions below

1. What are the dimensions of the dataframe? What do each of the numbers refer to?

```{r Dataframe Dimensions, eval = FALSE}

dim(gapminder)

```

2. What are the names of the columns in the data frame?

```{r Column Names, eval = FALSE}

names(gapminder)

```

3. What are the first and last countries in the data frame?

```{r Head and Tail of Data, eval = FALSE}

head(gapminder)

tail(gapminder)

```

4. What is the minimum and maximum gdpPercap? How many NAs are there for fertility? How many observations do we have for Africa?

```{r Data Summary, eval = FALSE}

summary(gapminder)

```

5. What years are covered in the data frame?

```{r Unique years, eval = FALSE}

unique(gapminder$year)

```

6. How many countries are in our data frame?

```{r Unique countries, eval = FALSE}

unique(gapminder$country)

```


## Extracting Information 

### Selecting a single value

If we wanted to get the first value from the first row and column in the dataframe we could use:

```{r Select the first element}

gapminder[1,1]

```


### Selecting a whole row

If we wanted the whole first row we could use:

```{r Select the first row}

gapminder[1,]

```

Notice, that if we want to select all columns we simply add a comma and leave the column position blank.

What happens if you run the following?

```{r The wrong way to select first row, eval = FALSE}

gapminder[1]

```

### Selecting specific rows and columns

If we wanted the first 5 rows and the first and third columns we could use:

```{r Select the first 5 rows and first and third columns}

gapminder[1:5, c(1,3)]

```

Remember from before that if we have nonconsecutive positions, we need to use the `c()` function to **c**ombine these positions into a list. 


### Reference columns by name

We can also reference the column by name:

```{r Referencing the column by column name}

gapminder[1:5, c("country", "gdpPercap")]

```

Why might this be preferred to referencing columns by number?


# Exercises Part III

1. Extract all rows from the column `pop` and save it in a new object called `pop`

Hint: look back to how we selected row 1 and all columns. 

```{r Select the column pop , eval = FALSE}

pop <- gapminder[]

```


2. Extract the 5th row and 6th column from the dataset

```{r Extract 5th row and 6th column, eval = FALSE}

gapminder[]

```


3. **Bonus** Extract all the rows for the columns `gdpPercap` and `pop`.

Hint: look back to how we selected row 1 and all columns. 


```{r Select columns gdpPercap and pop, eval = FALSE}

gapminder[]

```

4. Extract rows 5, 20, and 44 from the column `lifeExp` and save it in a new data frame called `sub_lifeExp`

```{r Extract rows 5 and 20 and 44 from col lifeExp A, eval = TRUE}

 sub_lifeExp <- gapminder[c(5, 20, 44), "lifeExp"]

```

